### 2016-11-15 greedy_hamiltonian_revised/chapters/5.numres.tex
fist commit

\subsection{Parametric Linear Wave equation} Consider the stochastic linear wave equation
\begin{equation} \label{eq:NuRe:1}
\left\{
\begin{aligned}
& u_{tt}(x,t,\omega) = \kappa(\omega) u_{xx}(x,t,\omega), \\
& u(x,0) = u^0(x),
\end{aligned}
\right.
\end{equation}
where $x$ belongs to a one-dimensional torus of length $L$, $\omega = (\omega_1,\dots,\omega_4)$ and
\begin{equation} \label{eq:NuRe:2}
	\kappa(\omega) = c^2\left( \sum_{l=1}^4 \frac{1}{l^2} \omega_l \right).
\end{equation}
Each $\omega_l$ is uniformly distributed in $[0,1]$ and $c\in \mathbb{R}$ is a constant number. By rewriting (\ref{eq:NuRe:1}) in canonical form, using the change of variable $q = u$ and $\partial q/ \partial t= p$, we obtain the symplectic form
\begin{equation} \label{eq:NuRe:3}
\left\{
\begin{aligned}
& q_t(x,t,\omega) = p(x,t,\omega), \\
& p_t(x,t,\omega) = \kappa(\omega) q_{xx}(x,t,\omega),
\end{aligned}
\right.
\end{equation}
with the associated Hamiltonian given by
\begin{equation} \label{eq:NuRe:4}
	H(q,p,\omega) = \frac 1 2 \int_0^L p^2 + \kappa(\omega) q_x^2 \ dx.
\end{equation}
We discretized the torus into $N$ equidistant points and define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\omega)$ and $p_i=p(t,x_i,\omega)$ for $i = 1, \dots, N$. Furthermore, we discretize (\ref{eq:NuRe:3}) using a standard central finite differences scheme to obtain
\begin{equation} \label{eq:NuRe:5}
	\frac{d}{dt} \mathbf z = \mathbb{J}_{2N} L\mathbf z,
\end{equation}
where $\mathbf z=(q,\dots,q_N,p_q,\dots,p_n)^T$ and
\begin{equation} \label{eq:NuRe:6}
L = 
\begin{pmatrix}
	I_n & 0_N \\
	0_N & \kappa(\omega)D_{xx}
\end{pmatrix},\quad 
\end{equation}
with $D_{xx}$ the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
\begin{equation} \label{eq:NuRe:7}
	H_{\Delta x}(\mathbf z) = \frac{\Delta x}2 \sum_{i=1}^{N} \left( p_i^2 + \kappa(\omega) \frac{(q_{i+1} - q_i)^2}{2\Delta x ^ 2} + \kappa(\omega) \frac{(q_{i} - q_{i-1})^2}{2\Delta x ^ 2} \right).
\end{equation}
The initial condition is given by
\begin{equation} \label{eq:NuRe:8}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:NuRe:9}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}
This will result in waves propagating in both directions on the torus.

For numerical time integration we used the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, which becomes an explicit scheme since the Hamiltonian is separable for the linear wave-equation. The full model uses the following parameter set \vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.01$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

We compare the reduced system obtained by the greedy algorithm with the methods based on SVD. To generate snapshots, we discretize the parameter space $[0,1]^4$ into in total of $5^4$ equidistant grid points. For the SVD based methods and POD, snapshots are gathered in the snapshot matrices $S$, $S_{\text{combined}}$ and $S_{\text{complex}}$, respectively, and the SVD is performed to construct the reduced basis. The greedy method is applied according to Algorithm \ref{alg:SyMo:3}; as input, the tolerance for the error in the Hamiltonian is set to $\delta = 5 \times 10^{-3}$. All reduced systems are taken to have an identical size ($k=80$ for POD and $k=40$ for the symplectic methods). We use the Verlet-Str\"omer scheme for symplectic methods and a second order Runge-Kutta method for the POD. Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.

@@

Figure \ref{fig:NuRe:2c} shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods. We note that the error for the POD reduced system rapidly increases, confirming that projection based reduced system does not yield a stable solution. Furthermore, the symplectic methods provide a better approximation since the geometric structure of the original system is preserved. Although the greedy method is almost twice faster than SVD-based method in the offline stage, its accuracy is comparable. For complex systems were the solution of the full system is expensive and for parameter domains with higher dimensions, we expect the greedy method to be superior in terms of computational cost.

@@

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:2a}\includegraphics[width=1\textwidth]{./figs/wave/singular}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:2b}\includegraphics[width=\textwidth]{./figs/wave/hamiltonian}}
\end{minipage}\par\medskip
\centering
\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}

@@

Regarding computation of the nonlinear terms of reduced systems, We compare the DEIM with the symplectic DEIM. For generation of the DEIM reduced basis we apply Algorithm \ref{alg:MoOr:1} to the set of nonlinear snapshots. Algorithm \ref{alg:SyMo:4} is used to construct a reduced basis appropriate for the symplectic DEIM. As input, we provide the symplectic basis generated by Algorithm \ref{alg:SyMo:3} with the set of nonlinear snapshots and a tolerance for the error $\delta = 10^{-4}$.

@@

\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=0$]{\label{fig:NuRe:3a}\includegraphics[width=1\textwidth]{./figs/schrodinger/solution/solution_t0}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=1$]{\label{fig:NuRe:3b}\includegraphics[width=\textwidth]{./figs/schrodinger/solution/solution_t10}}
\end{minipage}\par\medskip
\centering
\subfloat[$t=2$]{\label{fig:NuRe:3c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}}
\caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$. Here the solution of full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
\label{fig:NuRe:3}
\end{figure}

@@

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:4a}\includegraphics[width=1\textwidth]{./figs/schrodinger/singular}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:4b}\includegraphics[width=\textwidth]{./figs/schrodinger/hamiltonian}}
\end{minipage}\par\medskip
\centering
\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}

@@

The value of the Hamiltonian function for the reduced systems discussed above as a function of time is illustrated in Figure \ref{fig:NuRe:4b}. We see that the symplectic methods preserve the Hamiltonian function. We like to emphasise that the combination of the reduced basis by the greedy and the DEIM does not preserve the symplectic structure as suggested in this figure.

Figure \ref{fig:NuRe:4c} illustrates the $L^2$-error between the solution of the full model with the reduced systems generated by different methods. We first observe that symplectic methods yield a lower computational error compared to non-symplectic methods. Secondly, we observe that although the reduced systems from the cotangent lift and the complex SVD are of the same size, their accuracy is different by an order of magnitude. We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is less than 20\%. Lastly we notice that the combination of the greedy reduced basis and DEIM yields large errors in the solution while the solution using the symplectic DEIM is very accurate. We note that the symplectic DEIM is even more accurate than the greedy itself since it has been enriched by the nonlinear snapshots.

>>>>

\subsection{Parametric Linear Wave equation} Consider the {\edit parametric} linear wave equation
\begin{equation} \label{eq:NuRe:1}
\left\{
\begin{aligned}
& u_{tt}(x,t,\omega) = \kappa(\omega) u_{xx}(x,t,\omega), \\
& u(x,0) = u^0(x),
\end{aligned}
\right.
\end{equation}
where $x$ belongs to a one-dimensional torus of length $L$, $\omega = (\omega_1,\dots,\omega_4)$ and
\begin{equation} \label{eq:NuRe:2}
	\kappa(\omega) = c^2\left( \sum_{l=1}^4 \frac{1}{l^2} \omega_l \right).
\end{equation}
Each $\omega_l$ is uniformly distributed in $[0,1]$ and $c\in \mathbb{R}$ is a constant number. By rewriting (\ref{eq:NuRe:1}) in canonical form, using the change of variable $q = u$ and $\partial q/ \partial t= p$, we obtain the symplectic form
\begin{equation} \label{eq:NuRe:3}
\left\{
\begin{aligned}
& q_t(x,t,\omega) = p(x,t,\omega), \\
& p_t(x,t,\omega) = \kappa(\omega) q_{xx}(x,t,\omega),
\end{aligned}
\right.
\end{equation}
with the associated Hamiltonian given by
\begin{equation} \label{eq:NuRe:4}
	H(q,p,\omega) = \frac 1 2 \int_0^L p^2 + \kappa(\omega) q_x^2 \ dx.
\end{equation}
We discretized the torus into $N$ equidistant points and define $\Delta x = L/N$, $x_i = i\Delta x$, $q_i=q(t,x_i,\omega)$ and $p_i=p(t,x_i,\omega)$ for $i = 1, \dots, N$. Furthermore, we discretize (\ref{eq:NuRe:3}) using a standard central finite differences scheme to obtain
\begin{equation} \label{eq:NuRe:5}
	\frac{d}{dt} \mathbf z = \mathbb{J}_{2N} L\mathbf z,
\end{equation}
where $\mathbf z=(q,\dots,q_N,p_q,\dots,p_n)^T$ and
\begin{equation} \label{eq:NuRe:6}
L = 
\begin{pmatrix}
	I_n & 0_N \\
	0_N & \kappa(\omega)D_{xx}
\end{pmatrix},\quad 
\end{equation}
with $D_{xx}$ the central finite differences matrix operator. The discrete Hamiltonian can finally be written as
\begin{equation} \label{eq:NuRe:7}
	H_{\Delta x}(\mathbf z) = \frac{\Delta x}2 \sum_{i=1}^{N} \left( p_i^2 + \kappa(\omega) \frac{(q_{i+1} - q_i)^2}{2\Delta x ^ 2} + \kappa(\omega) \frac{(q_{i} - q_{i-1})^2}{2\Delta x ^ 2} \right).
\end{equation}
The initial condition is given by
\begin{equation} \label{eq:NuRe:8}
	q_i(0) = h( 10\times|x_i - \frac{1}{2}| ), \quad p_i = 0, \quad i=1,\dots,N
\end{equation}
where $h(s)$ is the cubic spline function
\begin{equation} \label{eq:NuRe:9}
h(s) = 
\left\{
\begin{aligned}
& 1 - \frac{3}{2}s^2 + \frac{3}{4}s^3, \quad & 0\leq s \leq 1, \\
& \frac{1}{4}(2-s)^3, & 1< s \leq 2, \\
& 0, & s > 2.
\end{aligned}
\right.
\end{equation}
This will result in waves propagating in both directions on the torus.

For numerical time integration we {\edit use} the Str\"omer-Verlet (\ref{eq:Hasy:13}) scheme, which becomes an explicit scheme since the Hamiltonian is separable for the linear wave-equation. The full model uses the following parameter set \vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 1$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = 0.002$ \\
Time discretization size & $\Delta t = 0.01$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

We compare the reduced system obtained by the greedy algorithm with the methods based on SVD. To generate snapshots, we discretize the parameter space $[0,1]^4$ into in total of $5^4$ equidistant grid points. For the SVD based methods and POD, snapshots are gathered in the snapshot matrices $S$, $S_{\text{combined}}$ and $S_{\text{complex}}$, respectively, and the SVD is performed to construct the reduced basis. The greedy method is applied according to Algorithm \ref{alg:SyMo:3}; as input, the tolerance for the error in the Hamiltonian is set to $\delta = 5 \times 10^{-3}$. All reduced systems are taken to have an identical size ($k=80$ for POD and $k=40$ for the symplectic methods). We use the Verlet-Str\"omer scheme for symplectic methods and a second order Runge-Kutta method for the POD. {\edit The choice of different time integration routines is due to the fact that POD destroys the canonical form of the original equations and a symplectic integrator cannot be applied. One can alternatively use separate reduced subspaces for the potential and the momentum spaces, which however is not a standard model approach and needs further analysis.} Finally we use transformation (\ref{eq:SyMo:1}) to transfer the solution of the reduced systems into the high-dimensional space for illustration purposes.

@@

Figure \ref{fig:NuRe:2c} shows the $L^2$-error between the solution of the full model and the reduced systems constructed by different methods. We note that the error for the POD reduced system rapidly increases, confirming that {\edit the} projection based reduced system does not yield a stable solution. Furthermore, the symplectic methods provide a better approximation since the geometric structure of the original system is preserved. Although the greedy method is almost {\edit two times faster than the SVD-based methods} in the offline stage, its accuracy is comparable. {\edit For complex systems were the solution of the full system is expensive and for high dimensional parameter domains, POD-based methods become exceedingly expensive \cite{Anonymous:2016wl,Quarteroni:2016wi}. However the greedy method requires substantially fewer (proportional to the size of the reduced basis) evaluation of the time integration of the original system. Therefore we expect the greedy method to be superior in terms of computational cost.}

@@

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=\textwidth]{./figs/wave/error}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:2b}\includegraphics[width=\textwidth]{./figs/wave/hamiltonian}}
\end{minipage}\par\medskip
\centering
%\subfloat[]{\label{fig:NuRe:2c}\includegraphics[width=0.5\textwidth]{./figs/wave/error}}

@@

Regarding computation of the nonlinear terms of reduced systems, {\edit we} compare the DEIM with the symplectic DEIM. For generation of the DEIM reduced basis we apply Algorithm \ref{alg:MoOr:1} to the set of nonlinear snapshots. Algorithm \ref{alg:SyMo:4} is used to construct a reduced basis appropriate for the symplectic DEIM. As input, we provide the symplectic basis generated by Algorithm \ref{alg:SyMo:3} with the set of nonlinear snapshots and a tolerance for the error $\delta = 10^{-4}$.

@@

\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=0$]{\label{fig:NuRe:3a}\includegraphics[width=1\textwidth]{./figs/schrodinger/solution/solution_t0}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[$t=1$]{\label{fig:NuRe:3b}\includegraphics[width=\textwidth]{./figs/schrodinger/solution/solution_t10}}
\end{minipage}\par\medskip
\centering
\subfloat[$t=2$]{\label{fig:NuRe:3c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/solution/solution_t20}}
\caption{The solution $|u(t,x)| = \sqrt{q^2 + p^2}$ at $t=0$, $t=10$ and $t=20$ of the Nonlinear Schr\"odinger equation for parameter value $\epsilon = 1.0932$. Here the solution of {\edit the} full system together with the solution of the POD, cotangent lift, complex SVD and the greedy reduced system is presented.}
\label{fig:NuRe:3}
\end{figure}

@@

\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=\textwidth]{./figs/schrodinger/error}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
\subfloat[]{\label{fig:NuRe:4b}\includegraphics[width=\textwidth]{./figs/schrodinger/hamiltonian}}
\end{minipage}\par\medskip
\centering
%\subfloat[]{\label{fig:NuRe:4c}\includegraphics[width=0.5\textwidth]{./figs/schrodinger/error}}

@@

{\edit In Figure \ref{fig:NuRe:4b} we observe that the symplectic methods preserve the Hamiltonian function, unlike the POD and the DIEM methods. We would like to emphasise that using the reduced basis by the greedy and together with the DEIM (purple line) does not preserve the symplectic structure as suggested in this figure.}

Figure \ref{fig:NuRe:4c} illustrates the $L^2$-error between the solution of the full model with the reduced systems generated by different methods. We first observe that symplectic methods yield a lower computational error compared to non-symplectic methods. Secondly, we observe that although the reduced systems from the cotangent lift and the complex SVD are of the same size, their accuracy is different by an order of magnitude. We notice that the greedy algorithm is slightly less accurate than the cotangent lift method while its offline computational cost is less than 20\% {\edit compared to the cotangent lift}. Lastly we notice that the combination of the greedy reduced basis and DEIM yields large errors in the solution while the solution using the symplectic DEIM is very accurate. We note that the symplectic DEIM is even more accurate than the greedy itself since it has been enriched by the nonlinear snapshots.

### 2016-11-15 greedy_hamiltonian_revised/chapters/6.conclusion.tex
fist commit

In this paper, we present a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems. An iteration of the greedy method comprises searching the parameter space using the error in the Hamiltonian, to find the best basis vectors that increase the overall accuracy of the reduced basis. We argued that for a compact subset with exponentially small Kolmogorov $n$-width we recover exponentially fast convergence of the greedy algorithm. For fast approximation of nonlinear terms, the basis obtained by the greedy was combined with the symplectic DEIM to construct a Hamiltonian reduced system.

>>>>

In this paper, we {\edit presented} a greedy approach for the construction of a reduced system that preserves the geometric structure of Hamiltonian systems. An iteration of the greedy method comprises searching the parameter space using the error in the Hamiltonian, to find the best basis vectors that increase the overall accuracy of the reduced basis. We argued that for a compact subset with exponentially small Kolmogorov $n$-width we recover exponentially fast convergence of the greedy algorithm. For fast approximation of nonlinear terms, the basis obtained by the greedy was combined with the symplectic DEIM to construct a Hamiltonian reduced system.

### 2016-11-15 greedy_hamiltonian_revised/paper.tex
fist commit

@@

% REQUIRED
\begin{abstract}
While reduced-order models (ROMs) are popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question. We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model. Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the reduce basis. We used the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors. We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters. Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure. This enables the reduction of computational cost for nonlinear Hamiltonian systems. The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.
\end{abstract}

@@

\appendix
\input{./chapters/appendix.tex}

>>>>

\newcommand{\edit}{\color{red}}

@@

% REQUIRED
\begin{abstract}
While reduced-order models (ROMs) {\edit has been} popular for approximately solving large systems of differential equations, the stability of reduced models over long-time integration remains an open question. We present a greedy approach for ROM generation of parametric Hamiltonian systems which captures the symplectic structure of Hamiltonian systems to ensure stability of the reduced model. Through the greedy selection of basis vectors, two new vectors are added at each iteration to the set of basis vectors to increase the overall accuracy of the {\edit reduced} basis. We {\edit use} the error in the Hamiltonian function due to model reduction, as an error indicator to search the parameter space and find the next best basis vectors. We show that the greedy algorithm converges with exponential rate, under natural assumptions on the set of all solutions of the Hamiltonian system under variation of the parameters. Moreover, we demonstrate that combining the greedy basis with the discrete empirical interpolation method also preserves the symplectic structure. This enables the reduction of computational cost for nonlinear Hamiltonian systems. The efficiency, accuracy, and stability of this model reduction technique is illustrated through simulations of the parametric wave equation and the parametric Schr\"odinger equation.
\end{abstract}

@@

%\appendix
%\input{./chapters/appendix.tex}
